Fast and scalable Gaussian process modeling with applications to astronomical time series and power spectrum estimation \vspace{-3\baselineskip}

### 2017-03-21 paper/ms.tex
started adding discussion of power spectrum estimation

\title{%
Fast and scalable Gaussian process modeling with applications
to astronomical time series
\vspace{-3\baselineskip}
}

@@

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that scale well with amount of data.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the application of this method by applying it to simulated and
real astronomical time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method is flexible, fast, and most importantly, interpretable.
There are a wide range of potential applications within astronomical data
analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

@@

Using \eq{celerite-psd}, we find that for a single \celerite\ term, this
requirement is met when
\begin{eqnarray}\eqlabel{psd-req}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2} > 0 \quad.
\end{eqnarray}
The denominator is positive for all $c_j \ne 0$ and it can be shown that, when
$c_j=0$, \eq{psd-req} is satisfied for all $\omega \ne d_j$ where the power is
identically zero.
Therefore, when $c_j \ne 0$, we require that the numerator is positive for all
$\omega$.
This requirement can also be written as
\begin{eqnarray}
a_j\,c_j > -b_j\,d_j \\
a_j\,c_j > b_j\,d_j \quad.
\end{eqnarray}
Furthermore, we can see that $a_j$ must be positive since $k(0) = a_j$ should
should be positive and, similarly, by requiring the covariance to be finite at
infinite lag, we obtain the constraint $c_j \ge 0$.
Without loss of generality, we can choose $d_j \ge 0$ and find the final
constraint:
\begin{eqnarray}
    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j|
\end{eqnarray}

%The general case with multiple \celerite\ terms can be checked using Sturm's
%theorem. The power spectrum is computed from taking the square of the Fourier
%transform of the data time series; hence, the power-spectrum must be non-negative.
%In constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
%then it is possible for the power spectrum to go negative, which violates the
%definition of the power spectrum.
%Consequently, if any of these coefficients is negative, it is necessary to
%check that the power spectrum is still non-negative.  Note that a non-negative
%power-spectrum does not require the auto-correlation
%function to be positive for all $\tau$.  %The positive power spectrum is
%related to the positive eigenvalues of the covariance matrix which are
%necessary for a positive-definite matrix in limiting cases
%\citep{Messerschmitt:2006}.  %We find empirically that requiring an everywhere
%positive power spectrum results in positive eigenvalues for the covariance
%matrix, and so we describe here how to ensure a positive power spectrum using
%Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_j\,c_j-b_j\,d_j\\
r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation which may be solved exactly for the
roots.

>>>>

\title{%
Fast and scalable Gaussian process modeling with applications
to astronomical time series and power spectrum estimation
\vspace{-3\baselineskip}
}

@@

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that scale well with amount of data.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the application of this method by applying it to simulated and
real astronomical time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.  The power spectrum of an irregularly sampled time series may be 
estimated with Bayesian inference, with a posterior estimate of the associated 
uncertainties, in principle avoiding the problems associated with window functions 
and aliasing.
The method is flexible, fast, and most importantly, interpretable.
There are a wide range of potential applications within astronomical data
analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat

@@

Using \eq{celerite-psd}, we find that for a single \celerite\ term, this
requirement is met when
\begin{eqnarray}\eqlabel{psd-req}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2} > 0 \quad.
\end{eqnarray}
The denominator is positive for all $c_j \ne 0$ and it can be shown that, when
$c_j=0$, \eq{psd-req} is satisfied for all $\omega \ne d_j$ where the power is
identically zero.
Therefore, when $c_j \ne 0$, we require that the numerator is positive for all
$\omega$.

%The general case with multiple \celerite\ terms can be checked using Sturm's
%theorem. The power spectrum is computed from taking the square of the Fourier
%transform of the data time series; hence, the power-spectrum must be non-negative.
%In constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
%then it is possible for the power spectrum to go negative, which violates the
%definition of the power spectrum.
%Consequently, if any of these coefficients is negative, it is necessary to
%check that the power spectrum is still non-negative.  Note that a non-negative
%power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  %The positive power spectrum is
%related to the positive eigenvalues of the covariance matrix which are
%necessary for a positive-definite matrix in limiting cases
%\citep{Messerschmitt:2006}.  %We find empirically that requiring an everywhere
%positive power spectrum results in positive eigenvalues for the covariance
%matrix, and so we describe here how to ensure a positive power spectrum using
%Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation which may be solved exactly for the
roots.
